---
title: "Week 1 Extra"
author: "Ehsan Pishva"
date: "21-4-2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

### **Downloads**
#### Data
[Click here to download the data (“inflammation-01.csv”)](https://www.dropbox.com/s/0h5rl8k6hx9se3v/inflammation-01.csv?dl=0)

#### R-script
[Click here to download the R-scripts used in week 1 extra](https://www.dropbox.com/s/9ahsuvxovlca5va/Self-Study_Week1%20Extra.R.r?dl=0)

---
<br>
<br>

### Exercise 1

These exercises all use the  data frame you can dowload from the link provided above ("inflammation-01.csv"). First we need to import it as a variable dat. The dataset contain a cytokine measurement for 40 consecutive days of in 60 individuals.  

```{r eval = FALSE}
setwd("C:/Users/e.pishva/Dropbox/Intro to R/data/")
#Change the working directory

dat <- read.csv(file = "inflammation-01.csv", header = TRUE, row.names = 1)
```


1. Select from the data frame 'dat':

    - The element on the 4th row and 2nd column.
    
    - The element on the 7th row and 5th column.
    
    - The 10th row.
    
    - The 3rd column.

<br>

2. Let’s calculate some summary statistics for specific individuals (rows) and days (columns).

    - Calculate the mean inflammation score for the first individual.
    
    - Calculate the median inflammataion score for the 5th day.
    
    - Use min() and max() to calculate the range of values for the 10th individual.
    
    - Check you answer with the range() function.
    
    - Use quantile() to calculate the [interquartile range] (https://en.wikipedia.org/wiki/Interquartile_range) of the 6th indiviudal.

<br>

3. Using the inflammation data frame inflammation-01.csv: Let’s pretend there was something wrong with the instrument on the first five days for every second patient (#2, 4, 6, etc.), which resulted in the measurements being twice as large as they should be.

    - Write a vector containing each affected patient (hint: ? seq)
    
    - Create a new data frame with in which you halve the first five days’ values in only those patients
    
    - Print out the corrected data frame to check that your code has fixed the problem


<br>
<br>


### **Solution**
Go to the link below if you want to compare your script to the solution.



[Solution script GitHub](https://github.com/ehsanpishva/IntroR-PSY4373/blob/main/Scripts/IntroR_solution_exercise_1.R) 



[Solution script RAW](https://raw.githubusercontent.com/ehsanpishva/IntroR-PSY4373/main/Scripts/IntroR_solution_exercise_1.R)